From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-15 dictionary.tex
Improvements to QCS for tori

@@

In this section we define an isomorphism
\[
S_G : \CSiso{G} \to \Hh^2(E^\bullet_G).
\]
Let $\cs{L} = (\gcs{L},\mu,\phi)$ be a character sheaf on $G$.
For each geometric point $x\in {\bar G}$, choose a basis $\{ v_x \}$ for $\gcs{L}_x$.
Through this choice, $\cs{L}$ determines functions
\begin{align*}
a : {\bar G}\times {\bar G} &\to \EEx & b : {\bar G} &\to \EEx \\
\mu_{x,y}(v_{x+y}) &= a(x,y) v_x \otimes v_y & \phi_x(v_{\Frob{G}(x)}) &= b(x) v_x.
\end{align*}
Condition~\ref{CS.2} implies that
\begin{equation}\label{2-cocyle}
a(x+y,z) a(x,y) = a(x,y+z) a(y,z)
\end{equation}
for all $x,y,z\in {\bar G}$, so $a \in Z^2({\bar G},\EEx)$.  Similarly, condition~\ref{CS.3} gives
\begin{equation}\label{nohom}
\frac{a(\Frob{G}(x),\Frob{G}(y))}{a(x,y)} =  \frac{b(x+y)}{b(x) b(y)}
\end{equation}
for $x, y \in {\bar G}$.
Let $\alpha \in C^0(\Weil{},C^2({\bar G},\EEx)$ be the $0$-cochain corresponding to $a$ and let $\beta\in C^1(\Weil{},C^1({\bar G},\EEx)$ be the cocycle such that $\beta(\Frob{})$ is $b$.  We will write both $\alpha$ and $\beta$ additively.
Then
\[
d_G\alpha =0, \qquad\qquad
d_{\Weil{}} \alpha = d_{G} \beta,\qquad\qquad
d_{\Weil{}} \beta =0;
\]
in other words,
\[\alpha\oplus \beta \in Z^2(E^\bullet_G).\]
Although the cocycle $\alpha\oplus \beta$ is not well defined by $\cs{L}$, its class in $\Hh^2(E^\bullet_G)$ is.
To see this, let $\{ v'_x \in \gcs{L}_x^\times \tq x \in {\bar G}\}$ be another choice and let $\alpha'\oplus \beta' \in Z^2(E^\bullet_G)$ be defined by $\cs{L}$ and this choice, as above.
Now let $\delta \in C^0(\Weil{},C^1({\bar G},\EEx))$ correspond to the function $d : {\bar G}\to \EEx$ defined by $v'_x = d(x) v_x$.
Chasing through \ref{CS.2} and \ref{CS.3}, we find
\[
\alpha'\oplus\beta' = \alpha\oplus\beta + d\delta,
\]
so the class $[\alpha\oplus\beta]$ of $\alpha\oplus\beta$ in $\Hh^2(E^\bullet_G)$ is independent of the choice made above. It is also easy to see that $[\alpha\oplus\beta] = [\alpha_0\oplus\beta_0]$ when $\cs{L} \iso \cs{L}_0$,
which concludes the definition of the function
\begin{align*}
S_G : \CSiso{G} &\to \Hh^2(E^\bullet_G)\\
[\cs{L}] &\mapsto [\alpha\oplus \beta].
\end{align*}
It is also easy to see that $[\alpha_1\oplus\beta_1] + [\alpha_2\oplus\beta_2] = [\alpha_3\oplus\beta_3]$ when $\cs{L}_3 = \cs{L}_1\otimes \cs{L}_2$, so $S_G$ is a group homomorphism.

@@

\begin{proof}
We have already seen, in Lemma~\ref{lem:autornaught}, that every $\rho\in \Hom(\cs{L},\cs{L}')$ is either trivial or an isomorphism.
Now suppose $\rho \in \Aut(\cs{L})$.
The second diagram in \ref{CS.4} shows that the association $x \mapsto \brho_x$ is a homomorphism from $G(\bFq)$ to $\EEx$ and the first diagram in \ref{CS.4} shows that it factors through $G(\bFq)_{\Weil{}}$.

Conversely, if $\rho : G(\bFq)_{\Weil{}} \to \EEx$ is any homomorphism, then defining $\brho_x$ as multiplication by $\rho(x)$ will define a morphism $\gcs{L} \to \gcs{L}'$ satisfy the two diagrams in \ref{CS.4}.

@@

\begin{theorem}\label{thm:autornaught}
Let $G$ be a smooth commutative group scheme over $\Fq$.
There is a canonical isomorphism
\[
\Aut(\cs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx).
\]
\end{theorem}
\begin{proof} 
Fix $\cs{L} = (\gcs{L},\mu,\phi)$ and consider the group homomorphism from $\Aut(\cs{L})$ to $\Hom(\bG_{\Weil{}},\EEx)$ defined in Proposition~\ref{prop:autornaught_etale}.
This homomorphism is injective because morphisms of sheaves are determined by the linear transformations induced on stalks.  
Homomorphisms in the image of $\Aut(\cs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ are continuous when $\bG$ is viewed as the base of the espace \'etal\'e attached to $\gcs{L}$.
Since $\ell$ is invertible in $\Fq$, it follows that the image of $\Aut(\cs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
We also have $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\cs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho^y: \gcs{L}^y \to \gcs{L}^y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gcs{L}\to \gcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\cs{L}\to \cs{L}$ which maps to $\theta$ under $\Aut(\cs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}

@@

\begin{theorem}\label{thm:CSXK}
Let $K$ be a local field with residue field $\Fq$, in which $\ell$ is invertible; 
let $R$ be the ring of integers of $K$.
\begin{enumerate}
\labitem{(1)}{x1}
The trace of Frobenius provides a natural transformation between the additive functors
\[
F_{\QCS}: X_K \mapsto \QCSiso{X}
\qquad\text{and}\qquad
F_{\text{QC}}: X_K \mapsto \Hom_{\text{QC}}(X(K),\EEx)
\]
as functors from $\mathcal{N}_K$ to the category of commutative groups.
\end{enumerate} 
Regarding this natural transformation, for every $X_K \in \mathcal{N}_K$ with N\'eron model $X$:
\begin{enumerate}[resume]
\labitem{(2)}{x2} there is a canonical short exact sequence of commutative groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Weil{}} \to \QCSiso{X_K} \to \Hom_\text{QC}(X(K),\EEx) \to 0;
\] 
\labitem{(3)}{x3} for all quasicharacter sheaves $\cs{F}$, $\cs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\cs{F},\cs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism;
\labitem{(4)}{x4} for all quasicharacter sheaves $\cs{F}$ for $X_K$, there is a canonical isomorphism
\[
\Aut(\cs{F}) \iso \Hom((\pi_0(X)_{\bFq})_{\Weil{}},\EEx).
\]
\end{enumerate}
\end{theorem}

\begin{proof}
%\begin{enumerate}
To prove \ref{x1}, use Proposition~\ref{prop:functorialG} with $G = \Gr^R_n(X)$, the fact that N\'eron models are unique up to isomorphism, the fact that every $\CS(\Gr^R_n(X))$ is a full subcategory of $\QCS(X)$ and the observation that every object in $\QCS(X)$ is in the essential image of $\CS(\Gr^R_n(X))$ for some $n$.
To prove \ref{x2}, use Theorem~\ref{thm:snake} with $G = \Gr^R_n(X)$ and then argue as in part \ref{x1}.
To prove \ref{x3}, argue as in the proof of Lemma~\ref{lem:autornaught}.
To prove \ref{x4}, use Theorem~\ref{thm:autornaught} with $G = \Gr^R_n(X)$, the fact that the component group of $\Gr^R_n(X)$ is independent of $n$ and then argue as in part \ref{x1}.
%\end{enumerate}
\end{proof}

@@

As we explained in the Introduction, our original motivation for this paper was to find a geometrization and categorification of quasicharacters of $p$-adic tori. 
This is is now provided by specializing Theorem~\ref{thm:CSXK} to the case when $X_K= T_K\in \mathcal{N}_K$ is an algebraic torus over $K$. 
Let $T$ be a Neron model for $T_K$.
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K},
\]
where $X_*(T_K)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Accordingly, by Theorem~\ref{thm:CSXK}, isomorphism classes of quasicharacter sheaves for $T_K$ determine quasicharacter sheaves according to the exact sequence
\[
0 \to \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{}} \to \QCSiso{T} \to \Hom_\text{CS}(T(K),\EEx) \to 0.
\]
Also by Theorem~\ref{thm:CSXK} and \cite{bitan:discriminant}*{Eq 3.1}, the automorphism group of every $\cs{F} \in \CS(T_K)$ is given by 
\[
\Aut(\cs{F}) \iso \Hom(X_*(T_K)_{\Weil{K}},\EEx),
\]
where $\Weil{K}$ is the Weil group for $K$. 
Since 
\[
\Hom(X_*(T_K)_{\Weil{K}},\EEx) \iso \Hom(X_*(T_K),\EEx)^{\Weil{K}}
\]
it follows that the automorphism group of every quasicharacter sheaf for $T$ is $(\check{T}_\ell)^{\Weil{K}}$, where $\check{T}_\ell$ is the $\ell$-adic dual torus to $T$.

>>>>

[Add reference to Takaski Suzuki's paper]

@@

In this section we define a group homomorphism
\[
S_G : \CSiso{G} \to \Hh^2(E^\bullet_G).
\]
Let $\cs{L} = (\gcs{L},\mu,\phi)$ be a character sheaf on $G$.
For each geometric point $x\in {\bar G}$, choose a basis $\{ v_x \}$ for $\gcs{L}_x$.
Through this choice, $\cs{L}$ determines functions
\begin{align*}
a : {\bar G}\times {\bar G} &\to \EEx & b : {\bar G} &\to \EEx \\
\mu_{x,y}(v_{x+y}) &= a(x,y) v_x \otimes v_y & \phi_x(v_{\Frob{G}(x)}) &= b(x) v_x.
\end{align*}
Condition~\ref{CS.2} implies that
\begin{equation}\label{2-cocyle}
a(x+y,z) a(x,y) = a(x,y+z) a(y,z)
\end{equation}
for all $x,y,z\in {\bar G}$, so $a \in Z^2({\bar G},\EEx)$.  Similarly, condition~\ref{CS.3} gives
\begin{equation}\label{nohom}
\frac{a(\Frob{G}(x),\Frob{G}(y))}{a(x,y)} =  \frac{b(x+y)}{b(x) b(y)}
\end{equation}
for $x, y \in {\bar G}$.
Let $\alpha \in C^0(\Weil{},C^2({\bar G},\EEx)$ be the $0$-cochain corresponding to $a$ and let $\beta\in C^1(\Weil{},C^1({\bar G},\EEx)$ be the cocycle such that $\beta(\Frob{})$ is $b$.  We will write both $\alpha$ and $\beta$ additively.
Then
\[
d_G\alpha =0, \qquad\qquad
d_{\Weil{}} \alpha = d_{G} \beta,\qquad\qquad
d_{\Weil{}} \beta =0;
\]
in other words,
\[\alpha\oplus \beta \in Z^2(E^\bullet_G).\]
Although the cocycle $\alpha\oplus \beta$ is not well defined by $\cs{L}$, its class in $\Hh^2(E^\bullet_G)$ is.
To see this, let $\{ v'_x \in \gcs{L}_x^\times \tq x \in {\bar G}\}$ be another choice and let $\alpha'\oplus \beta' \in Z^2(E^\bullet_G)$ be defined by $\cs{L}$ and this choice, as above.
Now let $\delta \in C^0(\Weil{},C^1({\bar G},\EEx))$ correspond to the function $d : {\bar G}\to \EEx$ defined by $v'_x = d(x) v_x$.
Chasing through \ref{CS.2} and \ref{CS.3}, we find
\[
\alpha'\oplus\beta' = \alpha\oplus\beta + d\delta,
\]
so the class $[\alpha\oplus\beta]$ of $\alpha\oplus\beta$ in $\Hh^2(E^\bullet_G)$ is independent of the choice made above. It is also easy to see that $[\alpha\oplus\beta] = [\alpha_0\oplus\beta_0]$ when $\cs{L} \iso \cs{L}_0$,
which concludes the definition of the function
\begin{align*}
S_G : \CSiso{G} &\to \Hh^2(E^\bullet_G)\\
[\cs{L}] &\mapsto [\alpha\oplus \beta].
\end{align*}
It is also easy to see that $[\alpha_1\oplus\beta_1] + [\alpha_2\oplus\beta_2] = [\alpha_3\oplus\beta_3]$ when $\cs{L}_3 = \cs{L}_1\otimes \cs{L}_2$, so $S_G$ is a group homomorphism.

@@

\begin{proof}
We have already seen, in Lemma~\ref{lem:autornaught}, that every $\rho\in \Hom(\cs{L},\cs{L}')$ is either trivial or an isomorphism.
Now suppose $\rho \in \Aut(\cs{L})$.
The second diagram in \ref{cs.4} shows that the association $x \mapsto \brho_x$ is a homomorphism from $G(\bFq)$ to $\EEx$ and the first diagram in \ref{cs.4} shows that it factors through $G(\bFq)_{\Weil{}}$.

Conversely, if $\rho : G(\bFq)_{\Weil{}} \to \EEx$ is any homomorphism, then defining $\brho_x$ as multiplication by $\rho(x)$ will define a morphism $\gcs{L} \to \gcs{L}'$ satisfy the two diagrams in \ref{cs.4}.

@@

\begin{theorem}\label{thm:autornaught}
Let $G$ be a smooth commutative group scheme over $\Fq$.
There is a canonical isomorphism
\[
\Aut(\cs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx).
\]
\end{theorem}
\begin{proof} 
Fix $\cs{L} = (\gcs{L},\mu,\phi)$ and consider the group homomorphism from $\Aut(\cs{L})$ to $\Hom(\bG_{\Weil{}},\EEx)$ defined in the proof of Proposition~\ref{prop:autornaught_etale}.
This homomorphism is injective because morphisms of sheaves are determined by the linear transformations induced on stalks.  
Homomorphisms in the image of $\Aut(\cs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ are continuous when $\bG$ is viewed as the base of the espace \'etal\'e attached to $\gcs{L}$.
Since $\ell$ is invertible in $\Fq$, it follows that the image of $\Aut(\cs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
We also have $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\cs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho^y: \gcs{L}^y \to \gcs{L}^y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gcs{L}\to \gcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\cs{L}\to \cs{L}$ which maps to $\theta$ under $\Aut(\cs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}

@@

\begin{theorem}\label{thm:CSXK}
Let $K$ be a local field with residue field $\Fq$, in which $\ell$ is invertible; 
let $R$ be the ring of integers of $K$.
\begin{enumerate}
\labitem{(0)}{x0}
The trace of Frobenius provides a natural transformation between the additive functors
\[
F_{\QCS}: X_K \mapsto \QCSiso{X}
\qquad\text{and}\qquad
F_{\text{QC}}: X_K \mapsto \Hom_{\text{QC}}(X(K),\EEx)
\]
as functors from $\mathcal{N}_K$ to the category of commutative groups.
\end{enumerate} 
Regarding this natural transformation, for every $X_K \in \mathcal{N}_K$ with N\'eron model $X$:
\begin{enumerate}[resume]
\labitem{(1)}{x1} there is a canonical short exact sequence of commutative groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Weil{}} \to \QCSiso{X} \to \Hom_\text{QC}(X(K),\EEx) \to 0;
\] 
\labitem{(2)}{x2} for all quasicharacter sheaves $\cs{F}$, $\cs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\cs{F},\cs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism;
\labitem{(3)}{x3} for all quasicharacter sheaves $\cs{F}$ for $X_K$, there is a canonical isomorphism
\[
\Aut(\cs{F}) \iso \Hom((\pi_0(X)_{\bFq})_{\Weil{}},\EEx).
\]
\end{enumerate}
\end{theorem}

\begin{proof}
%\begin{enumerate}
To prove \ref{x0}, use Proposition~\ref{prop:functorialG} with $G = \Gr^R_n(X)$, the fact that N\'eron models are unique up to isomorphism, the fact that every $\CS(\Gr^R_n(X))$ is a full subcategory of $\QCS(X)$ and the observation that every object in $\QCS(X)$ is in the essential image of $\CS(\Gr^R_n(X))$ for some $n$.
To prove \ref{x1}, use Theorem~\ref{thm:snake} with $G = \Gr^R_n(X)$ and then argue as in part \ref{x0}.
To prove \ref{x2}, argue as in the proof of Lemma~\ref{lem:autornaught}.
To prove \ref{x3}, use Theorem~\ref{thm:autornaught} with $G = \Gr^R_n(X)$, the fact that the component group of $\Gr^R_n(X)$ is independent of $n$ and then argue as in part \ref{x0}.
%\end{enumerate}
\end{proof}

@@

As we explained in the Introduction, our original motivation for this paper was to find a geometrization and categorification of quasicharacters of $p$-adic tori. 
This is now provided by specializing Theorem~\ref{thm:CSXK} to the case when $X_K\in \mathcal{N}_K$ is an algebraic torus over $K$.

\begin{corollary}
Let $T_K$ be an algebraic toris over $K$ and let $T$ be a Neron model for $T_K$.
The trace of Frobenius determines a canonical short exact sequence of commutative groups 
\[
0 \to \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{}} \to \QCSiso{T} \to \Hom_\text{QC}(T(K),\EEx) \to 0.
\] 
%For all quasicharacter sheaves $\cs{F}$, $\cs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\cs{F},\cs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
For all quasicharacter sheaves $\cs{F}$ for $X_K$, there is a canonical isomorphism
\[
\Aut(\cs{F}) \iso (\check{T}_\ell)^{\Weil{K}}.
\]
\end{corollary}

\begin{proof}
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K},
\]
where $X_*(T_K)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Accordingly, by Theorem~\ref{thm:CSXK}, the kernel of the trace of Frobenius is
\[
\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}} = \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{}}.
\] 
Also by Theorem~\ref{thm:CSXK}, $\Aut(\cs{F})  \iso \Hom((\pi_0(X)_{\bFq})_{\Weil{}},\EEx)$, canonically.
By \cite{bitan:discriminant}*{Eq 3.1} again, 
\[
\Hom((\pi_0(X)_{\bFq})_{\Weil{}},\EEx) \iso
\Hom(X_*(T_K)_{\Weil{K}},\EEx),
\]
canonically.
But $\Hom(X_*(T_K)_{\Weil{K}},\EEx) \iso \Hom(X_*(T_K),\EEx)^{\Weil{K}}$, canonically.
So, for any quasicharacter sheaf $\cs{F}$ for $T$,
\[
\Aut(\cs{F}) \iso (\check{T}_\ell)^{\Weil{K}},
\]
canonically, where $\check{T}_\ell$ is the $\ell$-adic dual torus $\Hom(X_*(T_K),\EEx)$.
\end{proof}
